Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-25 omrp.tex
streamlining the intro - II
1) Added a new reference
2) continuing on with the streamlining

\begin{abstract}
\noindent Overlapping Batch Means (OBM) has long been used in simulation as a method of reusing data to generate variance estimators with asymptotically lower variance.
In this paper, we extend the OBM method to stochastic programming by formulating a variant of the multiple replications procedure used for assessing solution quality.
We show conditions under which the resulting optimality gap estimators are strongly consistent, and have the same asymptotically lower variance as the classical results.
Computational experiments for several test problems are presented, examining the small-sample behavior of the overlapping batches method in this context.\smallskip

@@

Overlapping Batch Means (OBM) method, proposed by \citet{Meketon1984}, is widely used in simulation to obtain variance estimators with lower variances. 
In this paper we show that the same benefits can be obtained from overlapping batches when used in a stochastic optimization setting.  
A natural setting where batch means are used in stochastic optimization (esp. in stochastic programming) is for assessing solution quality. 
Assessing solution quality in stochastic programming typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because the optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators using the classical nonoverlapping batches method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the same variance reduction results can be achieved by overlapping like in the classical setting. 
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We present computational results to examine small-sample behavior. 
Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.

@@

Unless $f$ has a simple structure or the number of realizations of $\xi$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) is often generated by sampling from the distribution of $\xi$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the Sample Average Approximation (SAA), is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

We wish to assess the quality of a candidate solution $\xh \in X$. 
Assessing solution quality is important in practice because as (\ref{eq:sto_prog}) typically cannot be solved exactly, one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms designed to (approximately) solve (\ref{eq:sto_prog}).
We define the quality of $\xh$ by its optimality gap, $\e{f(\xh,\xit)} - \zs$. 
The lower the optimality gap, the higher the quality of the solution and a zero optimality gap implies that the solution is optimal. 
Note that the candidate solution $\xh \in X$ may have been found in any way. 
We assume this solution is given throughout this paper.

The optimality gap $\e{f(\xh,\xit)} - \zs$ cannot be calculated exactly. First,  for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ often involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian or semidefinite relations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator of optimality gap is then 
$$
G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

@@

In this paper, we extend MRP by overlapping the batches similar to the overlapping batch means (OBM) introduced by Meketon and Schmeiser \citep{Meketon1984}. 
As an example, suppose each batch contains $m$ observations and that there are a total of $2m$ observations.
In the nonoverlapping case, there are only two batches: the first batch contains observations $\xit^1, \xit^2, \dots, \xit^m$ and the second batch contains observations $\xit^{m+1}, \xit^{m+2}, \dots, \xit^{2m}$.   
Meketon and Schmeiser \citep{Meketon1984} overlap the batches so that the first batch is as before, the second batch now consists of $\xit^2, \xit^3, \dots, \xit^{m+1}$, the third batch contains observations $\xit^3, \xit^4, \dots, \xit^{m+2}$ and so on. 
As a result, the corresponding batch means are no longer independent. 
However, by simply reusing the data in this fashion, and seemingly in a contradictory way that induces dependence, one obtains a better variance estimator: a variance estimator that asymptotically has two-thirds of the variance of the usual batch means variance estimator but the bias of the two estimators is approximately the same.  
The idea of overlapping batches can be used for variance estimators of non-means \citep{SAH90} and recently, has been extended to estimators based on standardized time series (e.g., area, Cram\'{e}r-von Mises) for steady-state simulation output analysis \citep{Alexopoulos01012007,Alexopoulos2007}.

We explore the use of overlapping batches for assessing solution quality in stochastic programs.
Similar to Welch \citep{Welch1987} and Song and Schmeiser \citep{Song1992}, we consider partial overlap between batches to reduce computational effort in our setting.
The contributions of this paper are:
\begin{itemize}
	\item Formulate the overlapping multiple replications procedure for stochastic programs.
	\item Give conditions under which the point estimators are strongly consistent (that is, they converge almost surely (a.s.) as opposed in to in probability; referred to simply as consistent from this point on).
	\item Demonstrate that the classical variance reduction results are achieved for stochastic programs.
	\item Show the asymptotic validity of the resulting confidence intervals.
	\item Present computational results on two-stage stochastic linear and integer programs with recourse to examine small sample behavior.  
		Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.
\end{itemize}

@@

 We present variants of . .

@@

Batch means have been used in stochastic programming to assess solution quality and we show that. . . 
\bigskip

What happens when overlapping batches are used in a stochastic optimization setting?  Can we obtain the same benefits from overlapping batches as we do in a non-optimized setting?  In this paper, we answer the second question in the affirmative. We provide conditions under which the same benefits can be obtained by overlapping for a class of stochastic optimization problems. A natural setting where batch means is used in stochastic optimization is solution validation. This is referred to as the Multiple Replication Procedure, where each replication is essentially a batch means.

@@

Overlapping batch means modifies this idea by taking batches given by $\yb_j(m) = \frac{1}{m} \sum_{i=1}^{m} y^{(j-1)+i}$, for $j = 1, \dots, n-m+1$.  
This is what we call the \emph{maximally-overlapping} batch means method, because $m-1$ observations are common to---and only one observation changes between---adjacent batches.  
Given this, the sample variance estimator is updated to
\begin{equation} \label{eq:overlap_variance_def}
	\vart{\ybb} = \frac{1}{(\tfrac{n}{m}-1)(n - m + 1)}\sum_{j=1}^{n-m+1} (\yb_j(m) - \ybb)^2.
\end{equation}
Meketon and Schmeiser \citep{Meketon1984} show several attractive properties of the overlapping batches variance estimator, the two most important of which are: (i) the overlapping variance estimator has nearly the same bias as the standard nonoverlapping variance estimator, and (ii) the overlapping estimator has only two-thirds of the asymptotic variance of the nonoverlapping one.  
That is, 
\begin{equation} \label{eq:asym_var}
	\frac{ \var{\vart{\ybb}} }{ \var{\varhat(\ybb)} } \rightarrow \frac{2}{3}
\end{equation}
in the limit as batch size $m$ and then the number of batches $(n/m)$ tend to infinity.  
Note that the degrees of freedom in (\ref{eq:overlap_variance_def}) is slightly different than in \citep{Meketon1984}.  
The degrees of freedom in (\ref{eq:overlap_variance_def}) makes $\vart{\ybb}$ an unbiased estimator for iid data for all finite $m$ and $n$ with the same asymptotic benefits \citep{Song1992}.  
The $(1-\alpha)$-level approximate CI is similarly formed by $\ybb \pm t_{3(k-1)/2,\alpha/2} \sqrt{\vart{\ybb}}$, with an adjustment to the degrees of freedom.

@@

An advantage of MRP is its applicability to a wide range of problems.  
With iid sampling, (\ref{eq:sto_prog}) can be linear or nonlinear, $\X$ can include integrality constraints or not.  
It is also easy to implement, thus, has been applied to a variety of problems, see e.g., \citep{bertocchi_etal_99,janjarassuk_linderoth_08,santoso_ahmed_etal_05}.  
Recently, the approach of (nonoverlapping) batching has been used for assessing solution quality of stochastic programs with finitely many expected value \citep{wang_ahmed_08} and stochastic dominance \citep{hu2010sample} constraints.

@@

\begin{description}
	\item[A1] Samples of the random variable $\xit$ are iid.
	\item[A2] $\sup_{\x \in X} \e{f(\x,\xi)^{4}} < \infty$.
	\item[A3] (\ref{eq:sto_prog}) has a unique optimal solution $\xs$ and $\p{\xs_m \neq \xs} \rightarrow 0$ exponentially fast as $m \rightarrow \infty$.  
           That is, there exists constants $M > 0$ and $c > 0$ such that $\log(\p{\xs_m \neq \xs}) \leq -cm$ for all $m > M$.
\end{description}

First, we restrict our attention to iid sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied in several ways.  
Shapiro and Homem-de-Mello \citep{shapiro2000rate} establish rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite and the optimal solution is unique.   
Kleywegt, Shapiro,and Homem-de-Mello \citep{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition of a unique optimal solution $\xs$, and that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

\begin{proof}
	For brevity, we drop the $lj$ subscripts from $\db_{lj}$ and $\gb_{lj}$ and $l$ from $\gbb_l$ and $\dbb_l$.
	(i) We can decompose the expectation as
	\begin{align*}
		\e{(\db - \gb)^4} & = \e{\left[ \left.(\db - \gb)^4 \right| \db - \gb \neq 0\right]} \p{\db - \gb \neq 0} + 0\\
		& \leq \e{\left[ \left.(\db - \gb)^4 \right| \db - \gb \neq 0 \right]} \p{ \xs_{m_l}  \neq \xs}.
	\end{align*}
	The expectation term on the right-hand side is bounded by assumption A2 for iid samples.  
        Assumption A3 states $\p{\xs_{m_l} \neq \xs} \rightarrow 0$ exponentially fast.  
        Thus $\e{(\db - \gb)^4} \rightarrow 0$ exponentially fast.
(ii) This expectation term can be decomposed in the same manner
	\begin{align*}
		\e{(\dbb - \gbb)^4} & = \e{\left[ \left. (\dbb - \gbb)^4 \right| \dbb - \gbb \neq 0\right]} \p{\dbb - \gbb \neq 0}.
	\end{align*}
	With the expectation term on the right-hand side again bounded by A2, and the probability term converges to zero exponentially fast by Lemma \ref{lem:gbb_prob}.
\end{proof}

@@

	The fringe batches will come in identically distributed pairs.  
        The first and last batch would be identically distributed, as would the second and second-to-last, etc.  
        We know this because we could reverse the order of the iid sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		\left( \frac{n}{df} \right)^2 n_b \e{ \sum_{1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } & = \left( \frac{n}{df} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		& = \left( \frac{n}{df} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		& = \left( \frac{n}{df} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{df}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

>>>>

\begin{abstract}
\noindent Overlapping Batch Means (OBM) has long been used in simulation as a method of reusing data to generate variance estimators with asymptotically lower variance.
In this paper, we extend the OBM method to stochastic programming by formulating a variant of the multiple replications procedure used for assessing solution quality.
We show conditions under which the resulting optimality gap estimators are strongly consistent and have the same asymptotically lower variance as the classical results.
Computational experiments for several test problems are presented, examining the small-sample behavior of the overlapping batches method in this context.\smallskip

@@

Overlapping Batch Means (OBM) method, proposed by \citet{Meketon1984}, is widely used in simulation to obtain variance estimators with lower variances. 
While OBM is often used for variance estimators of means, it can also be used for variance estimators of non-means \citep{SAH90} and more recently, it has been extended to estimators based on standardized time series (e.g., area, Cram\'{e}r-von Mises) \citep{Alexopoulos01012007,Alexopoulos2007} and their folded versions \citep{Meterelliyoz_etal_12}.
In this paper we show that the same benefits can be obtained from overlapping batches when used in a stochastic optimization setting.  
A natural setting where batch means are used in stochastic optimization (esp. in stochastic programming) is for assessing solution quality. 
Assessing solution quality in stochastic programming typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because the optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators using the classical nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the same variance reduction results can be achieved by overlapping like in the classical setting. 
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We present computational results to examine small-sample behavior. 
Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.

@@

Unless $f$ has a simple structure or the number of realizations of $\xi$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) can be generated by sampling from the distribution of $\xi$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (i.i.d.) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the Sample Average Approximation (SAA), is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

We wish to assess the quality of a candidate solution $\xh \in X$. 
Assessing solution quality is important in practice because as (\ref{eq:sto_prog}) typically cannot be solved exactly, one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms designed to (approximately) solve (\ref{eq:sto_prog}).
We define the quality of $\xh$ by its optimality gap, $\e{f(\xh,\xit)} - \zs$. 
The lower the optimality gap, the higher the quality of the solution and a zero optimality gap implies that the solution is optimal. 
Note that the candidate solution $\xh \in X$ may have been found in any way. 
We assume that this solution is given as input to our procedures, and hence fixed, throughout this paper.

The optimality gap $\e{f(\xh,\xit)} - \zs$ often cannot be calculated exactly. First, for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ typically involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian or semidefinite relations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator of optimality gap is then 
$$
\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms.

@@

In this paper, we extend MRP by overlapping the batches similar to the overlapping batch means (OBM) introduced by Meketon and Schmeiser \citep{Meketon1984}. 
As an example, suppose each batch contains $m$ observations and that there are a total of $2m$ observations.
In the nonoverlapping case, there are only two batches: the first batch contains observations $\xit^1, \xit^2, \dots, \xit^m$ and the second batch contains observations $\xit^{m+1}, \xit^{m+2}, \dots, \xit^{2m}$.   
Meketon and Schmeiser \citep{Meketon1984} overlap the batches so that the first batch is as before, the second batch now consists of $\xit^2, \xit^3, \dots, \xit^{m+1}$, the third batch contains observations $\xit^3, \xit^4, \dots, \xit^{m+2}$ and so on. 
As a result, the corresponding batch means are no longer independent. 
However, by simply reusing the data in this fashion, and seemingly in a contradictory way that induces dependence, one obtains a better variance estimator: a variance estimator that asymptotically has two-thirds of the variance of the usual batch means variance estimator but the bias of the two estimators is approximately the same.

We explore the use of overlapping batches for assessing solution quality in stochastic programs.
Similar to Welch \citep{Welch1987} and Song and Schmeiser \citep{Song1992}, we consider partial overlap between batches to reduce computational effort in our setting.
%The contributions of this paper are:
%\begin{itemize}
%	\item Formulate the overlapping multiple replications procedure for stochastic programs.
%	\item Give conditions under which the point estimators are strongly consistent (that is, they converge almost surely (a.s.) as opposed in to in probability; referred to simply as consistent from this point on).
%	\item Demonstrate that the classical variance reduction results are achieved for stochastic programs.
%	\item Show the asymptotic validity of the resulting confidence intervals.
%	\item Present computational results on two-stage stochastic linear and integer programs with recourse to examine small sample behavior.  
%		Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.
%\end{itemize}

@@

% We present variants of . .

@@

%Batch means have been used in stochastic programming to assess solution quality and we show that. . . 
\bigskip

%What happens when overlapping batches are used in a stochastic optimization setting?  Can we obtain the same benefits from overlapping batches as we do in a non-optimized setting?  In this paper, we answer the second question in the affirmative. We provide conditions under which the same benefits can be obtained by overlapping for a class of stochastic optimization problems. A natural setting where batch means is used in stochastic optimization is solution validation. This is referred to as the Multiple Replication Procedure, where each replication is essentially a batch means.

%\bigskip

@@

Overlapping batch means modifies this idea by taking batches given by $\yb_j(m) = \frac{1}{m} \sum_{i=1}^{m} y^{(j-1)+i}$, for $j = 1, \dots, n-m+1$.  
This is what we call the \emph{maximally-overlapping} batch means method, because $m-1$ observations are common to---and only one observation changes between---adjacent batches.  
Given this, the sample variance estimator is updated to
\begin{equation} \label{eq:overlap_variance_def}
	\vart{\ybb} = \frac{1}{(\tfrac{n}{m}-1)(n - m + 1)}\sum_{j=1}^{n-m+1} (\yb_j(m) - \ybb)^2.
\end{equation}
Meketon and Schmeiser \citep{Meketon1984} show several attractive properties of the overlapping batches variance estimator, the two most important of which are: (i) the overlapping variance estimator has nearly the same bias as the standard nonoverlapping variance estimator, and (ii) the overlapping estimator has only two-thirds of the asymptotic variance of the nonoverlapping one.  
That is, 
\begin{equation} \label{eq:asym_var}
	\frac{ \var{\vart{\ybb}} }{ \var{\varhat(\ybb)} } \rightarrow \frac{2}{3}
\end{equation}
in the limit as batch size $m$ and then the number of batches $(n/m)$ tend to infinity.  
Note that the degrees of freedom in (\ref{eq:overlap_variance_def}) is slightly different than in \citep{Meketon1984}.  
The degrees of freedom in (\ref{eq:overlap_variance_def}) makes $\vart{\ybb}$ an unbiased estimator for i.i.d.\ data for all finite $m$ and $n$ with the same asymptotic benefits \citep{Song1992}.  
The $(1-\alpha)$-level approximate CI is similarly formed by $\ybb \pm t_{3(k-1)/2,\alpha/2} \sqrt{\vart{\ybb}}$, with an adjustment to the degrees of freedom.

@@

An advantage of MRP is its applicability to a wide range of problems.  
With i.i.d.\ sampling, (\ref{eq:sto_prog}) can be linear or nonlinear, $\X$ can include integrality constraints or not.  
It is also easy to implement, thus, has been applied to a variety of problems, see e.g., \citep{bertocchi_etal_99,janjarassuk_linderoth_08,santoso_ahmed_etal_05}.  
Recently, the approach of (nonoverlapping) batching has been used for assessing solution quality of stochastic programs with finitely many expected value \citep{wang_ahmed_08} and stochastic dominance \citep{hu2010sample} constraints.

@@

\begin{description}
	\item[A1] Samples of the random variable $\xit$ are i.i.d.
	\item[A2] $\sup_{\x \in X} \e{f(\x,\xi)^{4}} < \infty$.
	\item[A3] (\ref{eq:sto_prog}) has a unique optimal solution $\xs$ and $\p{\xs_m \neq \xs} \rightarrow 0$ exponentially fast as $m \rightarrow \infty$.  
           That is, there exists constants $M > 0$ and $c > 0$ such that $\log(\p{\xs_m \neq \xs}) \leq -cm$ for all $m > M$.
\end{description}

First, we restrict our attention to i.i.d.\ sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied in several ways.  
Shapiro and Homem-de-Mello \citep{shapiro2000rate} establish rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite and the optimal solution is unique.   
Kleywegt, Shapiro,and Homem-de-Mello \citep{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition of a unique optimal solution $\xs$, and that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

\begin{proof}
	For brevity, we drop the $lj$ subscripts from $\db_{lj}$ and $\gb_{lj}$ and $l$ from $\gbb_l$ and $\dbb_l$.
	(i) We can decompose the expectation as
	\begin{align*}
		\e{(\db - \gb)^4} & = \e{\left[ \left.(\db - \gb)^4 \right| \db - \gb \neq 0\right]} \p{\db - \gb \neq 0} + 0\\
		& \leq \e{\left[ \left.(\db - \gb)^4 \right| \db - \gb \neq 0 \right]} \p{ \xs_{m_l}  \neq \xs}.
	\end{align*}
	The expectation term on the right-hand side is bounded by assumption A2 for i.i.d.\ samples.  
        Assumption A3 states $\p{\xs_{m_l} \neq \xs} \rightarrow 0$ exponentially fast.  
        Thus $\e{(\db - \gb)^4} \rightarrow 0$ exponentially fast.
(ii) This expectation term can be decomposed in the same manner
	\begin{align*}
		\e{(\dbb - \gbb)^4} & = \e{\left[ \left. (\dbb - \gbb)^4 \right| \dbb - \gbb \neq 0\right]} \p{\dbb - \gbb \neq 0}.
	\end{align*}
	With the expectation term on the right-hand side again bounded by A2, and the probability term converges to zero exponentially fast by Lemma \ref{lem:gbb_prob}.
\end{proof}

@@

	The fringe batches will come in identically distributed pairs.  
        The first and last batch would be identically distributed, as would the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		\left( \frac{n}{df} \right)^2 n_b \e{ \sum_{1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } & = \left( \frac{n}{df} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		& = \left( \frac{n}{df} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		& = \left( \frac{n}{df} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{df} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{df}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}
